---
title: "Pronto_Exploration_2"
author: "Aaron Lichtner"
date: "November 6, 2015"
output: html_document
runtime: shiny
---
# How do people use the system? Cater to them.

```{r}
library(dplyr)
library(ggvis)
library(shiny)
library(knitr)
library(sp)
library(leaflet)
library(lubridate)
library(reshape2)
library(ggplot2)
library(DT)
library(shape)
opts_chunk$set(out.width='800px', dpi=200)
dtoptions = list(searching = FALSE, pageLength = 10, lengthChange = FALSE)
```


```{r}
df <- readRDS("data.rds")
```

# Net Elevation Change

```{r echo = FALSE}

m <- df %>% 
        group_by(Date) %>% 
        summarize(netElev = sum(elevation_diff, na.rm = TRUE)) %>% 
        ggvis(x = ~Date, y = ~netElev ) %>% 
        layer_points(size = 0.10, fill = "green") %>% 
        layer_smooths(span = 0.3, se = TRUE) %>% 
        add_axis("x", title = "", properties = axis_props(labels = list(angle = -50, align = "right"))) %>%
        add_axis("y", title = "Net Elevation Change", title_offset = 50) %>%
        add_legend(c("", "fill")) %>%
        bind_shiny("elevationDiff")

ggvisOutput("elevationDiff")
```

There was only one day (Day 1) where the system had a net positive elevation change. No doubt due to the number of early adopters trying out the system and challenging themselves to get up all of Seattle's hills. The trend quickly changed in how people use the system. 

### Net Elevation Change within the system normalized by the number of riders

```{r echo = FALSE}

m <- df %>% 
        group_by(Date) %>% 
        summarize(trips = n(), netElev = sum(elevation_diff, na.rm = TRUE))

scatter.smooth(m$Date, (m$netElev*3.28)/m$trips, span = 0.3, pch = 16, cex = 1.5, col = "#68D2DF", alpha = 0.5, axes = TRUE, lpars = c(col = "#003B49", lwd = 2))
abline(h = c(0, -20, -40, -60, -80), col = c("black", "grey", "grey", "grey", "grey"))

m %>% 
        ggvis(x = ~Date, y = ~netElev/trips) %>% 
        layer_points(size = 0.10, fill = "green") %>% 
        layer_smooths(span = 0.3, se = TRUE) %>% 
        add_axis("x", title = "", properties = axis_props(labels = list(angle = -50, align = "right"))) %>%
        add_axis("y", title = "Net Elevation Change", title_offset = 50) %>%
        add_legend(c("", "fill")) %>%
        bind_shiny("elevationDiffNorm")

ggvisOutput("elevationDiffNorm")

# Graph men and women's
```



```{r echo = FALSE}

m <- df %>% 
        group_by(Date, usertype) %>% 
        summarize(netElev = sum(elevation_diff, na.rm = TRUE)) %>% 
        ggvis(x = ~Date, y = ~netElev) %>% 
        layer_points(size = 0.10, fill = ~usertype) %>% 
        group_by(usertype) %>% 
        layer_smooths(span = 0.3, se = TRUE) %>%
        add_axis("x", title = "", properties = axis_props(labels = list(angle = -50, align = "right"))) %>%
        add_axis("y", title = "Normalized Net Elevation Change", title_offset = 50) %>%
        add_legend(c("fill")) %>%
        bind_shiny("elevUser")

ggvisOutput("elevUser")

```

### Net Elevation Change of by rider type normalized by number of riders

```{r echo = FALSE}

m <- df %>% 
        group_by(Date, usertype) %>% 
        summarize(trips = n(), netElev = sum(elevation_diff, na.rm = TRUE)) %>% 
        ggvis(x = ~Date, y = ~netElev/trips) %>% 
        layer_points(size = 0.10, fill = ~usertype) %>% 
        group_by(usertype) %>% 
        layer_smooths(span = 0.3, se = TRUE) %>%
        add_axis("x", title = "", properties = axis_props(labels = list(angle = -50, align = "right"))) %>%
        add_axis("y", title = "Normalized Net Elevation Change", title_offset = 50) %>%
        add_legend(c("fill")) %>%
        bind_shiny("elevUserNorm")

ggvisOutput("elevUserNorm")

```

```{r echo = FALSE}

m <- df %>% 
        filter(gender == "Male" | gender == "Female") %>%
        group_by(Date, gender) %>% 
        summarize(netElev = sum(elevation_diff, na.rm = TRUE)) %>% 
        ggvis(x = ~Date, y = ~netElev) %>% 
        layer_points(size = 0.10, fill = ~gender) %>%
        group_by(gender) %>% 
        layer_smooths(span = 0.3, se = TRUE) %>%
        add_axis("x", title = "", properties = axis_props(labels = list(angle = -50, align = "right"))) %>%
        add_axis("y", title = "Net Elevation Change", title_offset = 50) %>%
        add_legend(c("fill")) %>%
        bind_shiny("elevGender")

ggvisOutput("elevGender")

# Graph men and women's
```

### Gender separated elevation change normalized by number of riders

```{r echo = FALSE}
m <- df %>% 
        filter(gender == "Male" | gender == "Female") %>%
        group_by(Date, gender) %>% 
        summarize(trips = n(), netElev = sum(elevation_diff, na.rm = TRUE)) %>% 
        ggvis(x = ~Date, y = ~netElev/trips) %>% 
        layer_points(size = 0.10, fill = ~gender) %>% 
        group_by(gender) %>% 
        layer_smooths(span = 0.3, se = TRUE) %>%
        add_axis("x", title = "", properties = axis_props(labels = list(angle = -50, align = "right"))) %>%
        add_axis("y", title = "Normalized Net Elevation Change", title_offset = 50) %>%
        add_legend(c("fill")) %>%
        bind_shiny("elevGenderNorm")

ggvisOutput("elevGenderNorm")
```


```{r echo = FALSE}
m <- df %>% 
        group_by(Date, dayType) %>% 
        summarize(trips = n(), netElev = sum(elevation_diff, na.rm = TRUE)) %>% 
        ggvis(x = ~Date, y = ~netElev/trips) %>% 
        layer_points(size = 0.10, fill = ~dayType) %>% 
        group_by(dayType) %>% 
        layer_smooths(span = 0.3, se = TRUE) %>%
        add_axis("x", title = "", properties = axis_props(labels = list(angle = -50, align = "right"))) %>%
        add_axis("y", title = "Normalized Net Elevation Change", title_offset = 50) %>%
        add_legend(c("fill")) %>%
        bind_shiny("elevDayType")

ggvisOutput("elevDayType")
```


```{r echo = FALSE}
m <- df %>% 
        group_by(Date, Events) %>% 
        summarize(trips = n(), netElev = sum(elevation_diff, na.rm = TRUE)) %>% 
        ggvis(x = ~Date, y = ~netElev/trips) %>% 
        layer_points(size = 0.10, fill = ~Events) %>% 
        add_axis("x", title = "", properties = axis_props(labels = list(angle = -50, align = "right"))) %>%
        add_axis("y", title = "Normalized Net Elevation Change", title_offset = 50) %>%
        add_legend(c("fill")) %>%
        bind_shiny("elevWeather")

ggvisOutput("elevWeather")
```

```{r}
tripPairsFrom <- df %>% 
                group_by(from_station_name) %>% 
                summarize(lat = lat_from[1], long = long_from[1], numtrips = n()) %>% 
                arrange(desc(numtrips)) 

tripPairsTo <- df %>% 
                group_by(to_station_name) %>% 
                summarize(lat = lat_to[1], long = long_to[1], numtrips = n()) %>% 
                arrange(desc(numtrips)) 

# Trips from stations
par(mar = c(15,3,2,1))
barplot(tripPairsFrom$numtrips, names.arg = tripPairsFrom$from_station_name, cex.names = 0.50, las = 2)

tripPairsFrom %>% 
        leaflet() %>%
        addProviderTiles("CartoDB.Positron") %>% 
        setView(lng = -122.33, lat = 47.62, zoom = 12) %>% 
        addCircles(lng = ~long, lat = ~lat, weight = 1, radius = ~sqrt(numtrips) * 2, popup = ~from_station_name)
```


```{r}
# Destination Stations
par(mar = c(15,3,2,1))
barplot(tripPairsTo$numtrips, names.arg = tripPairsTo$to_station_name, cex.names = 0.50, las = 2)

        tripPairsTo %>% 
                leaflet() %>% 
                addProviderTiles("Stamen.Watercolor") %>% 
                setView(lng = -122.33, lat = 47.63, zoom = 12) %>% 
                addCircles(lng = ~long, lat = ~lat, weight = 1, radius = ~sqrt(numtrips) * 2, popup = ~to_station_name)
        
```


```{r}
# Show bikes coming vs. going for each station
        
        from <- df %>% filter(dayType == "Weekday" & from_station_name != "Pronto shop") %>%
                group_by(from_station_name, time = hour(starttime)) %>% 
                summarize(fromCount = n()) 
        
         toFrom <- df %>% filter(dayType == "Weekday" & to_station_name != "Pronto shop" & region_to == "FH" | region_to == "CD") %>%
                group_by(station = to_station_name, time = hour(stoptime)) %>% 
                summarize(toCount = n(), Area = region_to[1]) %>% 
                left_join(from, by = c("station" = "from_station_name", "time")) %>% 
                replace(is.na(.), 0) %>%
                mutate(netBikes = (toCount - fromCount)) %>% ungroup() %>%
                arrange(Area)
         
         wide <- dcast(toFrom, time ~ station)
         wide <- replace(wide, is.na(wide), 0)
         
         #68D2DF
         #003B49
         
         toFrom$station_f <- factor(toFrom$station, levels = unique(toFrom$station))
         
g <- ggplot()
g <- g + geom_area(data = toFrom, aes(time, toCount), fill = "blue") + 
         geom_area(data = toFrom, aes(time, -fromCount), fill = "darkgreen") + 
        coord_cartesian(ylim = c(-650, 700)) +
         facet_wrap(~station_f, ncol = 4) + 
         theme(axis.text = element_blank(), 
               axis.ticks = element_blank(), 
               plot.title = element_text(size = 15, vjust = 2),
               panel.grid.minor = element_blank(),
               panel.grid.major = element_blank(),
               panel.background = element_blank())
g

ggsave(filename = "ch.pdf", height = 6, units = "in", dpi = 300)

par(mfrow=c(7,8), mar=c(1,1,2,2))
for (i in 1:55) {
        plot(wide$time, wide[,(i+1)], type = "n")
        polygon(wide$time, wide[,(i+1)], col = "red")
}     

```        

### System wide usage by hour

```{r}
WeekdayHour <- df %>% filter(dayType == "Weekday") %>%
                group_by(time = hour(starttime)) %>% 
                summarize(count = n()/250) # Divide by number of weekdays (without holidays)

WeekendHour <- df %>% filter(dayType == "Weekend") %>%
                group_by(time = hour(starttime)) %>% 
                summarize(count = n()/114) # Divide by weekends and holidays

par(mfrow=c(2,1), mar=c(4,4,2,1))
        barplot(rep(NA,length(WeekendHour$count)),ylim=c(0,max(WeekdayHour$count)),axes=FALSE)
        abline(h=c(10, 20, 30, 40), col = "grey")
        barplot(WeekdayHour$count, col = "#003B49", axes = FALSE, add=T, border = FALSE)
        
        barplot(rep(NA,length(WeekendHour$count)),ylim=c(0,45),axes=FALSE)
        abline(h=c(10,20,30,40), col = "grey")
        barplot(WeekendHour$count, col = "#68D2DF", add = T, axes = FALSE, border = FALSE)
       

```

### How does it change by region?

```{r}
WeekdayRegionHour <- df %>% filter(dayType == "Weekday") %>%
                group_by(region = region_from, time = hour(starttime)) %>% 
                summarize(count = n()) 

WeekendRegionHour <- df %>% filter(dayType == "Weekend") %>%
                group_by(region = region_from, time = hour(starttime)) %>% 
                summarize(count = n()) 

wideWeekday <-  dcast(WeekdayRegionHour, time ~ region)
wideWeekday <- replace(wideWeekday, is.na(wideWeekday), 0)

par(mfrow=c(4,4), mar=c(3,1,2,2))
for (i in 1:13) {
        barplot(wideWeekday[,(i+1)], names.arg = colnames(wideWeekday[i+1]))
}

wideWeekend <-  dcast(WeekendRegionHour, time ~ region)
wideWeekend <- replace(wideWeekend, is.na(wideWeekend), 0)

par(mfrow=c(4,4), mar=c(3,1,2,2))
for (i in 1:13) {
        barplot(wideWeekend[,(i+1)], names.arg = colnames(wideWeekend[i+1]))
}


```

### Rides over the year (Weekday)

```{r}
rides <- df %>% filter(dayType == "Weekday") %>% 
        group_by(Date) %>% 
        summarize(count = n()) 

d <- ggplot(rides, aes(Date, count)) +
        geom_ribbon(aes(ymin = 0, ymax = count)) + 
        theme_bw()
d
```

### Annual vs. non-annual by station


### Temperature

```{r}

```

### Most common trip pairs

```{r}

commonPairs <- df %>% 
        group_by(from_station_name, to_station_name) %>% 
        summarize(count = n(), lat_from = mean(lat_from), lat_to = mean(lat_to), 
                  long_from = mean(long_from), long_to = mean(long_to)) %>% 
        arrange(desc(count)) %>% 
        filter(from_station_name != to_station_name) %>%
        top_n(count, n=1)

commonPairs

mostTrips <- df %>% group_by(from_station_name) %>% summarize(count = n(), lat_from = mean(lat_from), long_from = mean(long_from)) %>% arrange(desc(count))

most <- ggplot()

most <- most + geom_point(data=mostTrips, aes(x=long_from, y=lat_from, size=count), fill = "#8EDD65", color="#8EDD65", pch=21) + 
        geom_curve(data = commonPairs, aes(x = long_from, 
                                           y = lat_from,
                                           xend = long_to,
                                           yend = lat_to,
                                           size = 1), 
                                           color = "#8EDD65",
                                           lineend = "butt", 
                                           curvature = 0.4,
                                           angle = 90) +
        scale_size_continuous(range = c(1,20)) + theme_bw()
most

```

### Role of rain

```{r}

dailyPrec <- df %>% 
        filter(Precipitation_In != "NA") %>% 
        group_by(Date) %>% 
        summarize(count = n(), rain = mean(Precipitation_In)) %>% 
        arrange(desc(count)) %>% 
        group_by(rain) %>% summarize(avgRides = mean(count)) %>% 
        arrange(desc(avgRides)) 

m <- ggplot()
m <- m + geom_point(data = dailyPrec, aes(x = rain, y = avgRides), size = 3) + geom_smooth() + theme_bw()
m

dailyPrec%>%
        ggvis(x = ~rain, y = ~avgRides ) %>% 
        layer_points() %>% 
        layer_smooths(span = 0.3, se = TRUE) %>%
        add_axis("x", title = "Precipitation (inches)", properties = axis_props(labels = list(angle = -50, align = "right"))) %>%
        add_axis("y", title = "Average Trips", title_offset = 50) %>%
        bind_shiny("ridesRain")

ggvisOutput("ridesRain")


```

```{r}
dailyTemp <- df %>% 
        filter(Mean_Temperature_F != "NA") %>% 
        group_by(Date) %>% 
        summarize(count = n(), temp = mean(Mean_Temperature_F)) %>% 
        arrange(desc(count)) %>% 
        group_by(temp) %>% summarize(avgRides = mean(count)) %>% 
        arrange(desc(avgRides)) %>%
        ggvis(x = ~temp, y = ~avgRides ) %>% 
        layer_points() %>% 
        layer_smooths(span = 0.3, se = TRUE) %>%
        add_axis("x", title = "Temperature (F)", properties = axis_props(labels = list(angle = -50, align = "right"))) %>%
        add_axis("y", title = "Average Trips", title_offset = 50) %>%
        bind_shiny("ridesTemp")

ggvisOutput("ridesTemp")
```

There is a much stronger correlation between temperature and number of trips taken than daily precipition and trips taken.

```{r}
dailyWind <- df %>% 
        filter(Max_Wind_Speed_MPH != "NA") %>% 
        group_by(Date) %>% 
        summarize(count = n(), wind = mean(Max_Wind_Speed_MPH)) %>% 
        arrange(desc(count)) %>% 
        group_by(wind) %>% summarize(avgRides = mean(count)) %>% 
        arrange(desc(avgRides)) %>%
        ggvis(x = ~wind, y = ~avgRides ) %>% 
        layer_points() %>% 
        layer_smooths(span = 0.3, se = TRUE) %>%
        add_axis("x", title = "Wind Speed (MPH) ", properties = axis_props(labels = list(angle = -50, align = "right"))) %>%
        add_axis("y", title = "Average Trips", title_offset = 50) %>%
        bind_shiny("ridesWind")

ggvisOutput("ridesWind")
```

```{r}
dailyMeanWind <- df %>% 
        filter(Mean_Wind_Speed_MPH != "NA") %>% 
        group_by(Date) %>% 
        summarize(count = n(), windMean = mean(Mean_Wind_Speed_MPH)) %>% 
        arrange(desc(count)) %>% 
        group_by(windMean) %>% summarize(avgRides = mean(count)) %>% 
        arrange(desc(avgRides)) %>%
        ggvis(x = ~windMean, y = ~avgRides) %>% 
        layer_points() %>% 
        layer_smooths(se = TRUE) %>%
        add_axis("x", title = "Mean Wind Speed (MPH) ", properties = axis_props(labels = list(angle = -50, align = "right"))) %>%
        add_axis("y", title = "Average Trips", title_offset = 50) %>%
        bind_shiny("ridesMeanWind")

ggvisOutput("ridesMeanWind")
```

## Bikes

```{r}
trips <- df %>% group_by(bikeid) %>% 
        summarize(trips = n()) %>% 
        arrange(desc(trips)) %>% 
        ggvis(~trips) %>% 
        layer_histograms()

# histogram of trips per bike

#hist(trips$trips, breaks = 20, col = "#003B49", type = "n", main = "n")
#abline(h = c(20,40,60,80), col = "grey")
#hist(trips$trips, breaks = 20, col = "#003B49", border = FALSE, add = T)

# For people who stay within the 30 minute limit what is the average trip duration

df %>% 
        group_by(usertype) %>% filter(tripduration <= 3600) %>% ggvis(~tripduration/60) %>% layer_histograms(fill = ~usertype, stroke := "white", stack = FALSE)

df %>% 
        group_by(usertype) %>% filter(tripduration <= 3600) %>%
        summarize(duration = mean(tripduration, na.rm = TRUE)/60, sd = sd(tripduration, na.rm = TRUE)/60)



users <- df %>% group_by(usertype)
hist(users$tripduration, breaks = 30, freq = FALSE)

```


### Hills 

```{r}
hills <- df %>% group_by(elevation_diff) %>% summarize(count = n()) %>% arrange(desc(elevation_diff))

```